\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2017-06-12 eg6-letters/NIM/EG6_RTPC_NIM.tex
Implement Lamiaa's remarks

The original design of the RTPC~\cite{BONUS-NIM} was developed for the BoNuS 
experiment at Jefferson Lab which took data in CLAS in 2005. An enhanced design 
for use in the 2009 DVCS experiment is described in section \ref{sec_design} 
of this paper. Significant improvements were made to the RTPC mechanical 
structure and fabrication technique that both increased the acceptance and 
reduced the amount of material in the path of the outgoing particles. The 
data acquisition system described in section \ref{sec_readout} was reconfigured 
to increase the event readout rate. The calibration strategy and results for 
the RTPC are discussed in section \ref{sec_calib}. Finally, the performances
of the RTPC are described in section \ref{sec_perfor}.

@@

The CLAS RTPC is a 250~mm long cylinder of 158~mm diameter, leaving just enough 
room to fit pre-amplifiers between the RTPC outer shell and the solenoid. The 
electromagnetic field is directed perpendicularly to the beam direction, 
such that drifting electrons are pushed away from the beam line. These electrons 
are amplified by three layers of curved GEMs and detected by the readout system 
on the external shell of the detector as illustrated in the 
figure~\ref{fig:RTPC_1_4}. The CLAS RTPC is segmented into two halves with 
independent GEM amplification systems that cover about 80\% of the azimuthal 
angle.

We detail here the different regions shown in figure~\ref{fig:RTPC_1_4} starting 
from the beam line and towards larger radius:\\
\begin{itemize}
   \item First, the 6 atm helium-4 target extends along the beam line axis, it 
      is 284~mm long with a 3~mm radius wall made of a 27-$\mu$m-thick Kapton.
   \item The first gas gap covers a radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o ller electrons scattered by the beam. This 
      region is surrounded by a 4 $\mu$m thick window made of aluminized Mylar 
      is electronically grounded.
   \item The second gap region extends between 20~mm and 30~mm and is filled with a 
      gas mixture of 80$\%$ neon and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar, which 
      serves as the cathode and is set at 4260~V.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first GEM, 60 mm away 
      from the beam axis. The average electric field in this region is about 550~V/cm 
      and perpendicular to the beam.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied to each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Pre-amplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

@@

The GEM technology has been chosen for its flexibility as it
allows for a curved amplification surface. Also, GEMs are known to rarely create 
sparks~\cite{GEM_ref}, which is important when trying to detect 
highly ionizing slow nuclei that deposit large amount of energy. The GEMs for 
this RTPC are made from a Kapton insulator layer, 50~$\mu$m 
thick, sandwiched between two 5~$\mu$m copper layers. The mesh of each GEM 
layer is chemically etched with 50~$\mu$m diameter holes with double-conical 
shapes as illustrated in figure~\ref{fig:GEMs}. The potential difference 
applied between the two copper layers of the GEM creates a very strong 
electric field in each hole leading to high ionization and amplification.

The drift gas used in the experiment is a 80-20\% neon-DME mixture, this choice 
has been made in order to balance the energy deposit with a reasonable
Lorentz angle. Calculations using the MAGBOLTZ program~\cite{MAGBOLTZ} 
showed that with the perpendicular 5~T field, we would have a Lorentz angle of 
about 23$^\circ$ with this gas mixture. This is already a significant angle,
a heavier gas would have larger angle increasing significantly
the length of the drift paths as the electron would 
make a long spiral around the beam line before to reach the GEMs.

One of the important problematic to develop the radial TPC was to obtain a 
good support structure of the GEM foils, allowing an easy installation of the 
GEM in good conditions and with high precision. At the same time, we wanted to 
keep the material budget small in the forward region where we detect other 
particles in subsequent detectors. We successfully realized these prerequisites 
by using fiber glass rings glued to each ends of the GEM foils to form self 
supporting cylinders that could be installed independently in the RTPC after 
all gluing and soldering operations. The rigidity of the GEM foil is enough
for the structure to be self-supporting and only the backward part of the 
cylinder was fixed to the main mechanical structure. This design only left a 
light fiberglass ring in the forward region, reducing to a minimum secondary 
interactions.

@@

The RTPC electron collection system has 3200 readout pads. These elements are
located at the end of the amplification region, 69 mm from the central axis.
Figure \ref{fig:PADs} shows a schematic drawing of the size and 
configuration of the pads. Each readout pad is 5 mm in length and 4.45 mm in 
width.  The shift between the rows allows to reduce aliasing. Each half of the 
RTPC has 40 rows and 40 columns of pads. The shaded region in figure~\ref{fig:PADs} 
shows how pads are grouped to 16 channels pre-amplifiers. The pre-amplifier boards, 
already employed in the BoNuS RTPC~\cite{BONUS-NIM}, serve the dual purpose of 
inverting the RTPC signals polarity -- from negative to positive -- to match the 
requirements of the subsequent readout system, and driving the 6~m long ribbon 
cable that connects to it.

The readout system is based on the Front End Electronic (FEE) boards originally 
developed for the ALICE TPC readout system at CERN~\cite{ALICE-FEE}. Each 
readout channel is made of three main components: a charge sensitive, low 
impedance shaping amplifier, a 10 bit, 25-MHz digitizer, and a digital circuit 
that implements online-processing algorithms (pedestal subtraction, zero 
suppression, ad tail cancellation). The amplifier is implemented in a 
fully-analogue ASIC, PASA~\cite{ALICE-PASA}, while the subsequent signal 
digitization and elaboration part is implemented in a digital ASIC, 
ALTRO~\cite{ALICE-ALTRO}. Both ASICs host 16 channels. The full readout 
chain is implemented on a single Front End Card (FEC), hosting 128 channels. 
FECs are hosted in custom crates, mounted close to the detector. A Readout 
Control Unit (ROC) board is used to distribute the trigger signal to FECs 
and for data readout. Each RCU can handle up to 25 FECs. Communication 
between ROC and FECs is performed trough a custom back-plane, implementing 
a low-voltage signal bus. The RCU communicates with the CLAS DAQ system 
trough a 200 MB/s optical link, connected to a data acquisition PC hosting 
a ReadOut Receiver Card (RORC). This PC hosts the readout controller 
application (ROC) used to interface with the CLAS DAQ system. An Ethernet 
link is also present, for slow-controls and monitoring.

Triggering and data-readout scheme is as follows. When a Level-1 (L1) trigger 
is received by a FEC-board, a programmable number of samples $N_0$ is digitized 
and stored in a data memory, and hence processed by the digital circuit. Upon 
arrival of Level-2 trigger, the latest processed samples are saved in a 
multi-event memory, individual to each-channel, ready for readout by the RCU.  
In our setup, the RCU receives the trigger signal from the CLAS Trigger 
Supervisor system, and distributes both L1 and L2 signals to all the FECs in 
the crate, with a definite L1-L2 trigger latency. Such latency is programmable, 
and has to be long enough to accommodate the digitization of the $N_0$ samples 
per channel.  In the current implementation, $N_0=100$ samples/channel are 
acquired for each trigger, operating ALTRO at the reduced sampling frequency of 
$10 MHz$. The L1-L2 latency is thus fixed to 15 $\mu$s.

The trigger signal also initiates RCU readout operation from FEC boards. All 
the measured samples from active channels are reported, together with a channel 
identifier and a time-stamp, to the ROC application, that in turns sends them 
on the main event builder. During data reconstruction, the acquired samples are 
processed to obtain, for each readout pad, the accumulated charge (ADC) and the 
pulse time (T). Since pulse time was obtained as the time-stamp of the first 
sample above threshold, the resolution is equivalent to the ALTRO sampling 
time, 100~ns.

@@

In order to read all the detector readout pads, four FEC crates are used, 
each equipped with 6 boards, plus a ROC. A schematic of the readout system, 
for a single crate, is reported in figure~\ref{fig:ReadoutScheme}. This 
configuration permits to reduce the dead time associated to FEC readout 
operations, scaling linearly with the number of boards in the crate. During the 
2009 run, the system was successfully operated with a DAQ rate of 3.1 kHz, with 
a live time of $70 \%$.

@@

The timing information is used to infer the origin of the charge and 
then the trajectory of the detected particle
resulting in momentum measurement. Going from a collection of 
times to a momentum measurement requires a good knowledge of the drift speed 
and drift paths followed by the electrons released in the gas. The recorded ADCs give 
the deposited energy per unit of length ($\small{\frac{dE}{dX}}$) which, 
together with the momentum calculated from the trajectory, enables particle 
identification.

@@

We determine the drift speed using tracks reconstructed in the RTPC. In figure 
\ref{fig:RTPC_signals}, a typical $^{4}$He track is represented (in green). After 
it causes ionization in the drift region, the released electrons (in black) 
drift to the cylindrical detection plane under the effect of the electric field. The 
electrons released close to the cathode take the most time to reach the readout 
pads, but cylindrical symmetry insures they always travel the same 
distance. By identifying the maximum time measured ($T_{Max}$), we can infer the drift 
speed of the electrons in the RTPC.\\

@@

To measure the drift speed, we use the time profile of all hits in the 
chamber shown in figure \ref{fig:TDC_profile}. We can clearly observe the 
dropping edge expected from geometrical considerations. We define a value $T_{Max/2}$ at 
which the dropping edge passes half the maximum number of hits in the 
histogram. This value is measure in bins along the 200 mm RTPC's length to take into 
account variations in the electric and magnetic field in the RTPC (see figure 
\ref{fig:RunNumber_61551_TDCmax_Zslice}).

@@

Due to the non perfect experimental conditions, in particular variations in the 
gas mixture~\footnote{It was noted during the experiment that leaks appeared
and the gas flow had to be increased in few occasions to compensate for this.}, 
the drift speed changes during the three months long experimental run.  
Figure \ref{fig:Drift_run_number_1} shows the $T_{Max/2}$ values for 
individual runs (approximately 2 hours long). We observe a significant 
variation of the drift speed over time and accounted for it in the drift 
speed used for the track reconstruction.

@@

In summary, we obtain from our calibration a parametrization of the drift speed 
as a function of both position along the beam axis and run number (we did not 
observe any correlation between the effects). These 
functions were extracted for our entire data set and implemented in the 
track reconstructions code.

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. Software exists to calculate the drift paths, in particular
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used such calculation as a 
first calibration, but, 
as can be seen with the drift speed, gas composition is far from stable in the 
chamber. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.20 and 1.27 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematic allows to calculate the Helium nucleus kinematic. It is
by comparing this calculated kinematic to the measured one that we fine tune
the drift paths independently of our knowledge of the exact conditions in the 
chamber.

The drift paths are adjusted using a set of identified elastic events from our 
lower beam energy run. Based on the kinematic of the electrons in these events, 
we generate the helium nucleus in a GEANT4 simulation~\cite{GEANT4} of our 
RTPC. Then we compare the calculated GEANT4 trajectory of the Helium nuclei to 
the hits measured in the chamber.

@@

At the end of the extraction procedure, the azimuthal difference between the detection pad and 
the ionization point ($\Delta\phi$) is extracted as a function of time. 
In figure~\ref{fig:DELTA_PHI_TDC}, we show the resulting data points for one 
bin, where the drift paths is easily identified and eventually fitted for 
implementation in our reconstruction codes.

@@

To verify the stability of the drift paths, this procedure was carried out 
using both the 1.204 GeV data from the beginning of the run period and the 
1.269 GeV data from the end of the run period (shown in blue on 
figure~\ref{fig:Drift_run_number_1}). Interestingly, we found very similar drift paths
for the two data sets and concluded that any changes in the system only
significantly affected the drift speed.

@@

The first method is based on the measured average $\small{\frac{dE}{dX}}$
and the expected value based on the Bethe-Bloch formula. The total energy
deposition for each $^4$He track is calculated as the sum of all ADC hits
attributed to the track. The path length of that energy deposition is
calculated as the distance along the reconstructed trajectory between the first
and last hits on the track and corrected to account for regions
of bad channels along the track. The measured average energy loss is then the
ratio of deposited energy and path length. The Bethe-Bloch formula is
evaluated at the track's reconstructed momentum, and the ratio with the measured
$\small{\frac{dE}{dX}}$ is taken as a gain scaling factor. This average gain is
attributed equally to each readout channel contributing to the track, and each
channel's gain is averaged over many tracks. This gain calibration method is
inherently iterative, and, while it does result in measured $\small{\frac{dE}{dX}}$
much closer to the expected Bethe-Bloch curve than prior to calibration, the
results shown in figure~\ref{fig:dedx_p_exp_1st} are not very satisfactory.

@@

The second method is to compare the experimental ADCs to the energy deposited in 
GEANT4 by similar simulated tracks (using the same elastic events than for the drift 
paths calibration). This requires a very good GEANT4 simulation 
including drift paths, but also the spread of the charges along the path
before reaching the pad, so that the simulated hits match the experimental 
ones. Moreover, the simulation has to match the data acquisition (DAQ) features 
that can lead to cutting out hits. After setting the simulation properly, we 
compared simulation to experiment on an event by event basis as shown in 
Figure~\ref{fig:EVENT_adc_tdc}. In this step, the gain for each pad is calculated 
as the ratio of the measured ADCs to the simulated deposited energy.  
Then, these gains are refined using correcting factors obtained from a 
sample of good tracks. For each track, we calculate the corrected 
energy deposit on a pad and compare it to the average
deposit recorded by the other pads. 
Results of this second method are shown in Figure~\ref{fig:dedx_p_exp_2nd},
from these we concluded that the gains of the 
second method match best the theoretical lines.

@@

The first is an oscillatory noise located early in the readout time window, shown in the top panel of Figure~\ref{fig:noise} for a particularly noisy channel.  Its amplitude is not dissimilar to those of real tracks, and about 18\% of the readout channels exhibit large contributions from this noise characteristic.  Due to its unique time-energy correlation for the given channels, the noise could be removed on an event by event and channel by channel basis without significant loss of good signals, and the result is illustrated in the bottom panel of Figure~\ref{fig:noise}.

@@

\section{Track Reconstruction}\label{sec_rec}
In order to reconstruct tracks we first select good hits. This means rejecting 
out-of-time hits and hits linked to the electronic noise. The second step is 
to reconstruct the spacial origin of the hits using the extracted drift speed and drift path 
parameters. For each registered hit, we calculate a position of emission from 
the recorded time and the position of the recording pad. The third step 
is to create chains of hits. The maximum distance between two close adjacent 
hits has to be less than 10.5 mm to chain them, this roughly correspond to 
neighbors and next to neighbors. Then, we fit the chains that have a minimum of 
10 hits. We make the fit in two iterations, first, all the hits of the chain 
together with the beam line are fitted with a helix. For the second iteration, 
the hits that are 5 mm or farther from the first fit are excluded.

@@

The previous discussions have shown that the two modules of the RTPC have 
slightly different behavior and give also different particle yield. However, 
this should not necessarily be linked to a different performance of the RTPC. 
Indeed, there is a complicated convolution of CLAS and the RTPC acceptances 
that can significantly affect the yields and the kinematic of the detected 
particles. To clarify this question, we measured the 
efficiency of the RTPC using the elastic scattering on $^4$He, by comparing the 
 the inclusive yield based on electron detection only to the exclusive elastic 
yield where we also detect the Helium recoil. We present in 
Figure~\ref{fig:rtpc_eff} the results for the two modules 
of the detector. We observe that the left and 
the right modules have similar efficiencies except near the target windows.

>>>>

The original design of the RTPC~\cite{BONUS-NIM} was developed for the BoNuS 
experiment at Jefferson Lab which took data in CLAS in 2005. An enhanced design, 
used in the 2009 DVCS experiment, is described in section \ref{sec_design} 
of this paper. Significant improvements were made to the RTPC mechanical 
structure and fabrication technique that both increased the acceptance and 
reduced the amount of material in the path of the outgoing particles. The 
data acquisition system described in section \ref{sec_readout} was reconfigured 
to increase the event readout rate. The calibration strategy and results for 
the RTPC are discussed in section \ref{sec_calib}. Finally, the overall performance
of the RTPC is described in section \ref{sec_perfor}.

@@

The CLAS RTPC is a 250~mm long cylinder of 158~mm diameter, leaving just enough 
room to fit pre-amplifiers between the RTPC outer shell and the solenoid. The 
electromagnetic field is directed perpendicularly to the beam direction, 
such that drifting electrons are pushed away from the beam line. These electrons 
are amplified by three layers of curved GEMs and detected by the readout system 
on the external shell of the detector as illustrated in the 
Fig.~\ref{fig:RTPC_1_4}. The CLAS RTPC is segmented into two halves with 
independent GEM amplification systems that cover about 80\% of the azimuthal 
angle.

We detail here the different regions shown in Fig.~\ref{fig:RTPC_1_4} starting 
from the beam line towards larger radius:\\
\begin{itemize}
   \item First, the 6 atm helium-4 target extends along the beam line axis, it 
      is 284~mm long with a 3~mm radius wall made of a 27-$\mu$m-thick Kapton.
   \item The first gas gap covers a radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o ller electrons scattered by the beam. This 
      region is surrounded by a 4 $\mu$m thick window made of aluminized Mylar 
      that is electronically grounded.
   \item The second gap region extends between 20~mm and 30~mm and is filled with a 
      gas mixture of 80$\%$ neon (Ne) and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar, which 
      serves as the cathode and is set at 4260~V.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first GEM, 60 mm away 
      from the beam axis. The average electric field in this region is about 550~V/cm 
      and perpendicular to the beam.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied to each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Pre-amplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

@@

The GEM technology has been chosen for the flexibility of the GEM foils,
which can be easily used to produce a curved amplification surface. Also, 
GEMs are known to have relatively low sparks rate~\cite{GEM_ref}, which 
is important when trying to detect 
highly ionizing slow nuclei that deposit large amount of energy. The GEMs for 
this RTPC are made from a Kapton insulator layer, 50~$\mu$m 
thick, sandwiched between two 5~$\mu$m copper layers. The mesh of each GEM 
layer is chemically etched with 50~$\mu$m diameter holes with double-conical 
shapes as illustrated in Fig.~\ref{fig:GEMs}. The potential difference 
applied between the two copper layers of the GEM creates a very strong 
electric field in each hole leading to high ionization and amplification.

The drift gas used in the experiment is a 80-20\% Ne-DME mixture, this choice 
has been made in order to balance the energy deposit with a reasonable
Lorentz angle. Calculations using the MAGBOLTZ program~\cite{MAGBOLTZ} 
showed that with the perpendicular 5~T field, we would have a Lorentz angle of 
about 23$^\circ$ with this gas mixture. This is already a significant angle,
a heavier gas would have larger angle increasing significantly
the length of the drift paths as the electron would 
make a long spiral around the beam line to reach the GEMs.

One of the important challenge in developing the radial TPC was how to obtain a 
good support structure of the GEM foils to allow a tractable but highly precise 
installation of the GEMs. At the same time, we wanted to 
keep the material budget small in the forward region where we detect other 
particles in subsequent detectors. We successfully realized these prerequisites 
by using fiber glass rings glued to each ends of the GEM foils to form self 
supporting cylinders that could be installed independently in the RTPC after 
all gluing and soldering operations. The rigidity of the GEM foils was enough
for the structure to be self-supporting and only the backward part of the 
cylinder was fixed to the main mechanical structure. This design only left a 
light fiberglass ring in the forward region, reducing to a minimum secondary 
interactions.

@@

The RTPC electron collection system has 3200 readout pads. These elements are
located at the end of the amplification region, 69 mm from the central axis.
Figure~\ref{fig:PADs} shows a schematic drawing of the size and 
configuration of the pads. Each readout pad is 5 mm in length and 4.45 mm in 
width. The shift between the rows permits the reduction of aliasing. Each half of the 
RTPC has 40 rows and 40 columns of pads. The shaded region in Fig.~\ref{fig:PADs} 
shows how pads are grouped to 16 channels pre-amplifiers. The pre-amplifier boards, 
already employed in the BoNuS RTPC~\cite{BONUS-NIM}, serve the dual purpose of 
inverting the RTPC signals polarity -- from negative to positive -- to match the 
requirements of the subsequent readout system, and driv the 6~m long ribbon 
cable that connects to it.

The readout system is based on the Front End Electronic (FEE) boards originally 
developed for the ALICE TPC readout system at CERN~\cite{ALICE-FEE}. Each 
readout channel is made of three main components: a charge sensitive, low 
impedance shaping amplifier, a 10 bit, 25-MHz digitizer, and a digital circuit 
that implements online-processing algorithms (pedestal subtraction, zero 
suppression, and tail cancellation). The amplifier is implemented in a 
fully-analogue ASIC, PASA~\cite{ALICE-PASA}, while the subsequent signal 
digitization and elaboration part is implemented in a digital ASIC, 
ALTRO~\cite{ALICE-ALTRO}. Both ASICs host 16 channels. The full readout 
chain is implemented on a single Front End Card (FEC), hosting 128 channels. 
FECs are hosted in custom crates, mounted close to the detector. A readout 
control unit (RCU) board is used to distribute the trigger signal to FECs 
and for data readout. Each RCU can handle up to 25 FECs with communication 
between them performed trough a 
custom back-plane, implementing a low-voltage signal bus. The RCU communicates 
with the CLAS data acquisition (DAQ) system 
trough a 200 MB/s optical link, connected to a data acquisition PC hosting 
a ReadOut Receiver Card (RORC) used to interface with 
the CLAS DAQ system. An Ethernet link is also present, for slow-controls and 
monitoring.

Triggering and data-readout scheme is as follows. When a Level-1 (L1) trigger 
is received by a FEC-board, a programmable number of samples $N_0$ is digitized 
and stored in a data memory, and hence processed by the digital circuit. Upon 
arrival of Level-2 trigger, the latest processed samples are saved in a 
multi-event memory, individual to each-channel, ready for readout by the RCU.  
In our setup, the RCU receives the trigger signal from the CLAS Trigger 
Supervisor system, and distributes both L1 and L2 signals to all the FECs in 
the crate, with a definite L1-L2 trigger latency. Such latency is programmable, 
and has to be long enough to accommodate the digitization of the $N_0$ samples 
per channel.  In the current implementation, $N_0=100$ samples/channel are 
acquired for each trigger, operating ALTRO at the reduced sampling frequency of 
$10$~MHz. The L1-L2 latency is thus fixed to 15 $\mu$s.

The trigger signal also initiates RCU readout operation from FEC boards. All 
the measured samples from active channels are reported, together with a channel 
identifier and a time-stamp, to the ROC application, that in turns sends them 
on the main event builder. During data reconstruction, the acquired samples are 
processed to obtain, for each readout pad, the accumulated charge (ADC) and the 
pulse time (T). Since pulse time was obtained as the time-stamp of the first 
sample above threshold, the resolution is equivalent to the ALTRO sampling 
time of 100~ns.

@@

In order to read all the detector readout pads, four FEC crates are used, 
each equipped with 6 boards, plus a ROC. A schematic of the readout system, 
for a single crate, is reported in Fig.~\ref{fig:ReadoutScheme}. This 
configuration permits to reduce the dead time associated to FEC readout 
operations and to scale linearly with the number of boards in the crate. During the 
2009 run, the system was successfully operated with a DAQ rate of 3.1 kHz, with 
a live time of $70 \%$.

@@

The timing information is used to infer the origin of the charge and 
then the trajectory of the detected particle
resulting in momentum measurement. Going from a collection of 
times to a momentum measurement requires a good knowledge of the drift speed 
and drift paths followed by the electrons released in the gas. The recorded ADCs give 
the deposited energy per unit of length ($\small{\frac{dE}{dX}}$) which, 
together with the momentum calculated from the trajectory, enable the particle 
identification.

@@

We determine the drift speed using tracks reconstructed in the RTPC. In 
Fig.~\ref{fig:RTPC_signals}, a typical $^{4}$He track is represented (in green). After 
it causes ionization in the drift region, the released electrons (in black) 
drift to the cylindrical detection plane under the effect of the electric field. The 
electrons released close to the cathode take the most time to reach the readout 
pads, but cylindrical symmetry insures they always travel the same 
distance. By identifying the maximum time measured ($T_{Max}$), we can infer the drift 
speed of the electrons in the RTPC.\\

@@

To measure the drift speed, we use the time profile of all hits in the 
chamber shown in Fig.~\ref{fig:TDC_profile}. We can clearly observe the 
dropping edge expected from geometrical considerations. We define a value $T_{Max/2}$ at 
which the dropping edge passes half the maximum number of hits in the 
histogram. This value was measured in bins along the 200 mm RTPC's length to take into 
account variations in the electric and magnetic field in the RTPC (see 
Fig.~\ref{fig:RunNumber_61551_TDCmax_Zslice}).

@@

Due to the non perfect experimental conditions, in particular variations in the 
gas mixture~\footnote{It was noted during the experiment that leaks appeared
and the gas flow had to be increased in few occasions to compensate for this.}, 
the drift speed changes during the three months long experimental run.  
Figure~\ref{fig:Drift_run_number_1} shows the $T_{Max/2}$ values for 
individual runs (approximately 2 hours long). We observe a significant 
variation of the drift speed over time and accounted for it in the drift 
speed used for the track reconstruction.

@@

In summary, we obtain from our calibration a parametrization of the drift speed 
as a function of both position along the beam axis and the run number. These 
functions were extracted for our entire data set and implemented in the 
track reconstructions code.

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. Software exists to calculate the drift paths, in particular
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used such calculation as a 
first calibration, but, 
as can be seen with the drift speed, gas composition is far from stable in the 
chamber. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.20 and 1.27 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematic allows to calculate directly the Helium nucleus one. 
By comparing this calculated kinematic to the measurement in the RTPC, we fine tuned
the drift paths independently of our knowledge of the exact chamber's conditions.

The drift paths are adjusted using a set of identified elastic events from our 
lower beam energy run. Based on the kinematic of electrons in these events, 
we generate the helium nucleus in our RTPC GEANT4 simulation~\cite{GEANT4}. Then 
we compare the calculated GEANT4 trajectory of the Helium nuclei to 
the hits measured in the chamber.

@@

At the end of the extraction procedure, the azimuthal difference between the detection pad and 
the ionization point ($\Delta\phi$) is extracted as a function of time. 
In Fig.~\ref{fig:DELTA_PHI_TDC}, we show the resulting data points for one 
bin, where the drift paths is easily identified and eventually fitted for 
implementation in our reconstruction codes.

@@

To verify the stability of the drift paths, this procedure was carried out 
using both the 1.204 GeV data from the beginning of the run period and the 
1.269 GeV data from the end of the run period (shown in blue on 
Fig.~\ref{fig:Drift_run_number_1}). Interestingly, we found very similar drift paths
for the two data sets and concluded that any changes in the system only
significantly affected the drift speed.

@@

The first method is based on the measured average $\small{\frac{dE}{dX}}$
and the expected value based on the Bethe-Bloch formula. The total energy
deposition for each $^4$He track is calculated as the sum of all ADC hits
attributed to the track. The path length of that energy deposition is
calculated as the distance along the reconstructed trajectory between the first
and last hits on the track and corrected to account for regions
of bad channels along the track. The measured average energy loss is then the
ratio of deposited energy and path length. The Bethe-Bloch formula is
evaluated at the track's reconstructed momentum, and the ratio with the measured
$\small{\frac{dE}{dX}}$ is taken as a gain scaling factor. This average gain is
attributed equally to each readout channel contributing to the track, and each
channel's gain is averaged over many tracks. This gain calibration method is
inherently iterative, and, while it does result in measured $\small{\frac{dE}{dX}}$
much closer to the expected Bethe-Bloch curve than prior to calibration, the
results shown in Fig.~\ref{fig:dedx_p_exp_1st} are not very satisfactory.

@@

The second method is to compare the experimental ADCs to the energy deposited in 
GEANT4 by similar simulated tracks (using the same elastic events than for the drift 
paths calibration). This requires a very good GEANT4 simulation 
including drift paths, but also the spread of the charges along the path
before reaching the pad, so that the simulated hits match the experimental 
ones. Moreover, the simulation has to match the DAQ features 
that can lead to cutting out hits. After setting the simulation properly, we 
compared simulation to experiment on an event by event basis as shown in 
Fig.~\ref{fig:EVENT_adc_tdc}. In this step, the gain for each pad is calculated 
as the ratio of the measured ADCs to the simulated deposited energy.  
Then, these gains are refined using correcting factors obtained from a 
sample of good tracks. For each track, we calculate the corrected 
energy deposit on a pad and compare it to the average
deposit recorded by the other pads. 
Results of this second method are shown in Fig.~\ref{fig:dedx_p_exp_2nd},
from these we concluded that the gains of the 
second method match best the theoretical lines.

@@

The first is an oscillatory noise located early in the readout time window, 
shown in the top panel of Fig.~\ref{fig:noise} for a particularly noisy 
channel. Its amplitude is not dissimilar to those of real tracks, and about 
18\% of the readout channels exhibit large contributions from this noise 
characteristic. Due to its unique time-energy correlation for the given 
channels, the noise could be removed on an event by event and channel by 
channel basis without significant loss of good signals. The result of the
procedure is illustrated in the bottom panel of Fig.~\ref{fig:noise}.

@@

\section{Track Reconstruction}\label{sec_rec}
In order to reconstruct tracks we first select good hits. This means rejecting 
out-of-time hits as well as the hits linked to the electronic noise. The second step is 
to reconstruct the spacial origin of the hits using the extracted drift speed and drift path 
parameters. For each registered hit, we calculate a position of emission from 
the recorded time and the position of the recording pad. The third step 
is to create chains of hits. The maximum distance between two close adjacent 
hits has to be less than 10.5 mm to chain them, which roughly corresponds to 
neighbors and next to neighbors. Then, we fit with a helix the chains that have a minimum of 
10 hits. We make the fit in two iterations, first with all the hits of the chain 
and the beam line, then, in the second iteration, 
the hits that are 5 mm or farther from the first fit are excluded.

@@

The previous discussions have shown that the two modules of the RTPC have 
slightly different behavior and give also different particle yield. However, 
this should not necessarily be linked to a different performance of the RTPC. 
Indeed, there is a complicated convolution of CLAS and the RTPC acceptances 
that can significantly affect the yields and the kinematic of the detected 
particles. To clarify this question, we measured the 
efficiency of the RTPC using the elastic scattering on $^4$He, by comparing the 
inclusive yield based only on electron detection to the exclusive elastic 
yield where the Helium recoil is also detected. We present in 
Fig.~\ref{fig:rtpc_eff} the results for the two modules 
of the detector. We observe that the left and 
the right modules have similar efficiencies except near the target windows.
